Decomposing biodiversity change to processes of extinction, colonization, and persistence across scales

### 2023-04-19 scripts/ms_scaling_changes.qmd
initial commit

---
title: "Decomposing biodiversity change to processes of extinction, colonization, and persistence across scales"
author: "François Leroy et al."
format: 
  html:
    toc: true
    toc_float: true
editor: visual
---

@@

```{r}
## Create the dataframe to predict
nCEP_topredict <-
atlas_data_topredict %>% 
  # filter(!(((KVADRAT == 6844 & Date == 2015 & effort != 5))|(KVADRAT == 7248 & Date == 2015 & effort != 5))) %>% 
  filter(AREA != 0) %>% 
  dplyr::select(-c(Date, time_span, sr, effort)) %>% 
  distinct() %>% 
  mutate(effort_diff = 0,
         timespan_diff = 0) %>% 
  ## add the periods
  crossing(period = c("M2M3", "M3M4", "M2M4"))

@@

@@

pdf("../figures/fig4_sr.pdf",

@@

  # geom_violin(aes(area, slope, group = interaction(area, period)))+

@@

  # stat_summary(aes(area, slope, color = period, shape = period), fun.y = median, geom = 'point', size = 2.5)+
  # ylim(-2, 2)+

@@

@@

```{r}
pdf("../figures/col_ext_2periods.pdf",
    height = 5, width = 5.5)

for(THE_metric in c("col", "ext")){
print(predicted_nCEP_atlas %>% dplyr::select(spatial_scale, contains("M", ignore.case = F)) %>%
        dplyr::rename(AREA = spatial_scale) %>% 
        rbind(predicted_nCEP_BBS %>% dplyr::select(contains("M", ignore.case = F), AREA) %>% 
                mutate(AREA = ifelse(AREA < 0.2, 0.13, 2.51))) %>% 
        dplyr::select(AREA, contains(THE_metric)) %>% 
        pivot_longer(cols = contains(THE_metric),
                     names_to = "period",
                     values_to = "value") %>% 
        mutate(period = str_remove(period, THE_metric)) %>% 
          mutate(period = case_when(
            period == "M2M4" ~ "1985-2017",
            period == "M2M3" ~ "1985-2003",
            period == "M3M4" ~ "2003-2017")) %>% 
        mutate(period = fct_relevel(period, c("1985-2017", "1985-2003", "2003-2017"))) %>%
        filter(period != "1985-2017") %>%
        ggplot()+
        geom_boxplot(aes(AREA, value, group = interaction(AREA, period), color = period))+
        stat_summary(aes(AREA, value, color = period), fun.y = median, geom = 'line', size = 1, linetype = 2)+
        scale_x_log10()+
        annotation_logticks(sides = "b")+
        xlab("Spatial grain (km²)")+
        ylab(THE_metric)+
        ylim(0,50)+
        labs(color = "Period", shape = "Period", linetype = "Period")+
        theme_bw()+
        theme(text = element_text(size = 16),
              legend.position = "none"))
}
dev.off()
```

@@

>>>>

---
title: "Decomposing biodiversity change to processes of extinction, colonization, and persistence across scales"
author: "François Leroy et al."
format: 
  html:
    toc: true
    toc_float: true
    chunks:
    warning: false
    message: false
editor: visual
---

@@

```{r}
## Create the dataframe to predict
nCEP_topredict <-
atlas_data_topredict %>% 
  filter(AREA != 0) %>% 
  dplyr::select(-c(Date, time_span, sr, effort)) %>% 
  distinct() %>% 
  mutate(effort_diff = 0,
         timespan_diff = 0) %>% 
  ## add the periods
  crossing(period = c("M2M3", "M3M4", "M2M4"))

@@

## Probability of Colonization, Extinction and Persistence

```{r}
pCEP_atlas_models <- readRDS("../models/probaCEP_atlas.rds")
```

```{r}
## Create the dataframe to predict
pCEP_topredict <-
atlas_data_topredict %>% 
  filter(!(((KVADRAT == 6844 & Date == 2015 & effort != 5))|(KVADRAT == 7248 & Date == 2015 & effort != 5))) %>% 
  filter(AREA != 0) %>% 
  dplyr::select(-c(Date, time_span, sr, effort)) %>% 
  distinct() %>% 
  mutate(effort_diff = 0,
         timespan_diff = 0) %>% 
  ## add the periods
  crossing(period = c("M2M3", "M3M4", "M2M4"))

## And make predictions for each metric
predicted_pCEP_atlas <- data.frame()
## predictions for each metric
for(THE_metric in names(pCEP_atlas_models)){
  model <- pCEP_atlas_models[[THE_metric]][["xgboost"]]
  pred <- predict(model, 
                  newdata = pCEP_topredict)
  predicted_pCEP_atlas <- 
  rbind(predicted_pCEP_atlas,
        pCEP_topredict %>% cbind(pred) %>% mutate(metric = THE_metric))
}
```

```{r}
pCEP_BBS_models <- readRDS("../models/probaCEP_JPSP.rds")
```

```{r}
predicted_pCEP_BBS <- data.frame()
for(THE_metric in names(pCEP_BBS_models)){
  model <- pCEP_BBS_models[[THE_metric]][["rf"]]
  pred <- 
    predict(model, 
            newdata = JPSP_topredict %>% 
              dplyr::select(-time_span, -id, -sr, -Date, polypoint_ratio, Lat, Long, elevation, AREA) %>% 
              mutate(timespan_diff = 0) %>% 
              crossing(period = c("M2M3", "M3M4", "M2M4")))
  predicted_pCEP_BBS <- 
  rbind(
    predicted_pCEP_BBS,
    JPSP_topredict %>% 
    dplyr::select(time_span, Lat, Long, elevation, AREA) %>% 
    mutate(metric = THE_metric) %>% 
    crossing(period = c("M2M3", "M3M4", "M2M4")) %>% 
    cbind(pred)
  )
}

@@

pdf("../figures/sr_2periods.pdf",

@@

@@

@@

```{r}
pdf("../figures/trend_details.pdf", 
    height = 3, width = 11.297)
## Plot the detailed trends
JPSP_topredict %>% 
  mutate(AREA = ifelse(AREA < 0.2, 0.13, 2.51)) %>% 
  dplyr::select(sr, Date, AREA) %>% 
  rbind(
    pred130sqkm %>% dplyr::select(pred_sr, Date) %>% mutate(AREA = 130) %>% dplyr::rename(sr = pred_sr),
    pred530sqkm %>% dplyr::select(pred_sr, Date) %>% mutate(AREA = 530) %>% dplyr::rename(sr = pred_sr),
    pred2150sqkm %>% dplyr::select(pred_sr, Date) %>% mutate(AREA = 2150) %>% dplyr::rename(sr = pred_sr),
    predCR %>% dplyr::select(pred_sr, Date) %>% mutate(AREA = 83900) %>% dplyr::rename(sr = pred_sr)
  ) %>% 
  ggplot(aes(Date, sr))+
  stat_summary(fun = median, 
               geom = "point", col = "red", size = 2)+
  theme_bw()+
  stat_summary(aes(group = 1), fun.y = median,
               geom = "line", linetype = "dashed", size = 1, color = "blue")+
  facet_wrap(~ AREA, nrow = 1)+
  ylab("Species richness")+
  xlab("Year")+
  scale_x_continuous(breaks = c(1990,2000,2010))+
  theme(
    strip.background = element_blank(),
    strip.text.x = element_blank(),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 16))
dev.off()
```

@@

```{r}
pdf("../figures/col_ext_jaccard_2periods.pdf",
    height = 5, width = 5.5)
for(THE_metric in c("jaccard", "col", "ext")){
  if(THE_metric == "jaccard"){
    print(predicted_nCEP_atlas %>% dplyr::select(spatial_scale, contains("M", ignore.case = F)) %>% 
            dplyr::rename(AREA = spatial_scale) %>% 
            rbind(predicted_nCEP_BBS %>% 
                    dplyr::select(contains("M", ignore.case = F), AREA) %>% 
                    mutate(AREA = ifelse(AREA < 0.2, 0.13, 2.51))) %>% 
            dplyr::select(AREA, contains(THE_metric)) %>% 
            pivot_longer(cols = contains(THE_metric),
                         names_to = "period",
                         values_to = "value") %>% 
            mutate(period = str_remove(period, THE_metric)) %>% 
            mutate(period = case_when(
              period == "M2M4" ~ "1985-2017",
              period == "M2M3" ~ "1985-2003",
              period == "M3M4" ~ "2003-2017")) %>% 
            mutate(period = fct_relevel(period, c("1985-2017", "1985-2003", "2003-2017"))) %>%
            filter(period != "1985-2017") %>%
            ggplot()+
            geom_boxplot(aes(AREA, value, group = interaction(AREA, period), color = period))+
            stat_summary(aes(AREA, value, color = period), fun.y = median, geom = 'line', size = 1, linetype = 2)+
            scale_x_log10()+
            annotation_logticks(sides = "b")+
            xlab("Spatial grain (km²)")+
            ylab(THE_metric)+
            labs(color = "Period", shape = "Period", linetype = "Period")+
            theme_bw()+
            theme(text = element_text(size = 16),legend.position = "none"))
  }else{
    print(predicted_nCEP_atlas %>% dplyr::select(spatial_scale, contains("M", ignore.case = F)) %>%
        dplyr::rename(AREA = spatial_scale) %>% 
        rbind(predicted_nCEP_BBS %>% dplyr::select(contains("M", ignore.case = F), AREA) %>% 
                mutate(AREA = ifelse(AREA < 0.2, 0.13, 2.51))) %>% 
        dplyr::select(AREA, contains(THE_metric)) %>% 
        pivot_longer(cols = contains(THE_metric),
                     names_to = "period",
                     values_to = "value") %>% 
        mutate(period = str_remove(period, THE_metric)) %>% 
          mutate(period = case_when(
            period == "M2M4" ~ "1985-2017",
            period == "M2M3" ~ "1985-2003",
            period == "M3M4" ~ "2003-2017")) %>% 
        mutate(period = fct_relevel(period, c("1985-2017", "1985-2003", "2003-2017"))) %>%
        filter(period != "1985-2017") %>%
        ggplot()+
        geom_boxplot(aes(AREA, value, group = interaction(AREA, period), color = period))+
        stat_summary(aes(AREA, value, color = period), fun.y = median, geom = 'line', size = 1, linetype = 2)+
        scale_x_log10()+
        annotation_logticks(sides = "b")+
        xlab("Spatial grain (km²)")+
        ylab(THE_metric)+
        ylim(0,50)+
        labs(color = "Period", shape = "Period", linetype = "Period")+
        theme_bw()+
        theme(text = element_text(size = 16),
              legend.position = "none"))
  }
}
dev.off()
```

@@

## Probability of Colonization, Extinction, Persistence

```{r}
pdf("../figures/pCEP.pdf",
    height = 5, width = 5.5)

for(time in c("M2M3", "M3M4", "M2M4")) {
  print(
  predicted_pCEP_BBS %>%
  mutate(spatial_scale = ifelse(AREA < 0.2, 0.13, 2.51)) %>%
  mutate(period = fct_relevel(period, c("M2M4", "M2M3", "M3M4"))) %>%
  dplyr::select(spatial_scale, pred, period, metric) %>%
  rbind(predicted_pCEP_atlas %>% dplyr::select(spatial_scale, pred, period, metric)) %>%
  mutate(metric = case_when(
    metric == "col" ~ "Colonization",
    metric == "ext" ~ "Extinction",
    metric == "pers" ~ "Persistence"
  )) %>%
  filter(period == time,
         metric != "Species pool\ncolonization") %>%
  ggplot()+
  geom_boxplot(aes(spatial_scale, pred,
                   group = interaction(spatial_scale,metric),
                   fill = metric),
                   outlier.shape = NA) +
  scale_fill_brewer(palette = "Dark2")+
  scale_color_brewer(palette = "Dark2")+
  scale_x_log10()+
  annotation_logticks(sides = "b")+
  stat_summary(aes(spatial_scale, pred, color = metric), fun.y = median, geom = 'line', linetype = 2, size = 1)+
  ylim(0,1)+
  ylab("Probability")+
  xlab("Spatial grain\n(km²)")+
  labs(fill = "Probability")+
  theme_bw()+
  theme(text = element_text(size = 16),
        legend.position = "none"))
}
dev.off()
```

```{r}
pdf("../figures/pCEP_2periods.pdf",
    height = 5, width = 5.5)
for(THE_metric in c("col", "ext")){
  print(predicted_pCEP_BBS %>% 
  mutate(spatial_scale = ifelse(AREA < 0.2, 0.13, 2.51)) %>%
  mutate(period = fct_relevel(period, c("M2M4", "M2M3", "M3M4"))) %>% 
  dplyr::select(spatial_scale, pred, period, metric) %>% 
  rbind(predicted_pCEP_atlas %>% dplyr::select(spatial_scale, pred, period, metric)) %>% 
  filter(pred >= 0 & pred <= 1) %>% 
  filter(period != "M2M4") %>%
  filter(metric == THE_metric) %>%
  ggplot()+
  geom_boxplot(aes(spatial_scale, pred, group = interaction(spatial_scale, period), color = period))+
  stat_summary(aes(spatial_scale, pred, color = period, linetype = period), fun.y = median, geom = 'line', linetype = 2, size = 1)+
  scale_x_log10()+
  annotation_logticks(sides = "b")+
  xlab("Spatial grain (km²)")+
  ylab(THE_metric)+
  ylim(0, .5)+
  labs(color = "Period", shape = "Period", linetype = "Period")+
  theme_bw()+
  theme(text = element_text(size = 16))+
  theme(panel.background = element_blank(),
        legend.position = "none",
        panel.grid.major = element_line(colour = "lightgrey"),
        axis.text.x = element_markdown()))
}
dev.off()
```
